---
title: "inflexion stuff"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

date: 6 October 2022

```{r prep, message=FALSE, warning=FALSE, echo = FALSE}
library(magrittr) # assignment pipe
library(tidyverse) # default awesome functions
library(lubridate) # time manipulations
library(readxl) # alternative excel reader

# for beautiful plots
library(grid) # for grob
library(ggpubr) # from gridded plots

library(bookdown) # rmarkdown helpers
library(knitr)

library(glue) # alternative to paste

library(mgcv)
```

```{r get all data, echo = FALSE, warning=FALSE}
cases_national <- read_csv("https://raw.githubusercontent.com/ESR-NZ/covid_in_wastewater/main/data/cases_national.csv") %>%
  janitor::clean_names()
ww_national <- read_csv("https://raw.githubusercontent.com/ESR-NZ/covid_in_wastewater/main/data/ww_national.csv") %>% 
  janitor::clean_names() %>% 
  rename(copies_per_person_day = copies_per_day_per_person)
National_weekly <- left_join(ww_national, cases_national, 
                             by = "week_end_date")%>% 
  filter(week_end_date > ymd(20220101)) %>% 
  mutate(case_sum = case_7d_avg * 7)
National_weekly <-data.frame(x = rownames(National_weekly),
                             National_weekly,
                             row.names=NULL) %>% 
  arrange(week_end_date) %>% 
  mutate(x = as.numeric(x), 
         copies_last_week = lag(copies_per_person_day), 
         cases_last_week = lag(case_sum))
```

## EDA
```{r EDA, fig.width = 12, fig.height=12}
National_weekly %>% 
  # select(-x, -national_pop, -cases_per_100k, -week_epi) %>% 
  GGally::ggpairs() +
  theme_bw()
```

## BORING LM

```{r lm BORING}
model_wkcase_lm <- lm(data = National_weekly,
                      log10(case_sum) ~ cases_last_week + log10(copies_per_person_day) + log10(copies_last_week) + x)

par(mfrow = c(2,2))
plot(model_wkcase_lm)
summary(model_wkcase_lm)

National_weekly[8,]

National_weekly$y_hatcase_lm <- 10^predict(model_wkcase_lm, list(x = National_weekly$x, 
                                                                 copies_per_person_day = National_weekly$copies_per_person_day, 
                                                                 cases_last_week = National_weekly$cases_last_week, 
                                                                 copies_last_week = National_weekly$copies_last_week))
ggplot(data = National_weekly)+
  geom_line(aes(x = week_end_date, y = case_sum))+
  geom_line(aes(x = week_end_date, y = y_hatcase_lm), colour = "brown", size = 1.5)+
  scale_y_log10(labels = scales::label_log())+
  scale_x_date(labels = scales::label_date_short(),
               date_breaks = "2 week",
               name = "")+
  # geom_vline(xintercept = National_weekly[10,"week_end_date"])+
  theme_classic()
```

## POLYNOMIAL LM
```{r lm polynomial}
model_wkcase_lm1 <- lm(data = National_weekly,
                       log10(case_sum) ~ cases_last_week + log10(copies_per_person_day) + log10(copies_last_week) + x +
                         I(cases_last_week^2) + I(log10(copies_per_person_day)^2) + I(log10(copies_last_week)^2) + I(x^2))
# par(mfrow = c(2,2))
# plot(model_wkcase_lm1)
summary(model_wkcase_lm1)


model_wkcase_lm1 <- lm(data = National_weekly,
                       log10(case_sum) ~ cases_last_week + copies_per_person_day + copies_last_week + x +
                         I(cases_last_week^2) + I(log10(copies_per_person_day)^2) + I(log10(copies_last_week)^2) )
summary(model_wkcase_lm1)

National_weekly$y_hatcase_lm1 <- 10^predict(model_wkcase_lm1, list(x = National_weekly$x, 
                                                                   copies_per_person_day = National_weekly$copies_per_person_day,
                                                                   copies_last_week = National_weekly$copies_last_week,
                                                                   cases_last_week = National_weekly$cases_last_week,
                                                                   x = National_weekly$x))
ggplot(data = National_weekly)+
  geom_line(aes(x = week_end_date, y = case_sum))+
  geom_line(aes(x = week_end_date, y = y_hatcase_lm), colour = "brown")+
  geom_line(aes(x = week_end_date, y = y_hatcase_lm1), colour = "green", size = 1.5)+
  scale_y_log10(labels = scales::label_log())+
  scale_x_date(labels = scales::label_date_short(),
               date_breaks = "2 week",
               name = "")+
  theme_classic()
```


## GAM

```{r GAM}
model_wkcase_gam <- gam(data = National_weekly,
                        log10(case_sum) ~ s(cases_last_week) + s(log10(copies_per_person_day)) + s(log10(copies_last_week)) + s(x, k = 2))

par(mfrow = c(2,2))
plot(model_wkcase_gam)
summary(model_wkcase_gam)

National_weekly$y_hatcasegam <- 10^predict(model_wkcase_gam, list(x = National_weekly$x, 
                                                                  copies_per_person_day = National_weekly$copies_per_person_day,
                                                                  copies_last_week = National_weekly$copies_last_week,
                                                                  cases_last_week = National_weekly$cases_last_week,
                                                                  x = National_weekly$x))

ggplot(data = National_weekly)+
  geom_line(aes(x = week_end_date, y = y_hatcase_lm), colour = "brown")+
  geom_line(aes(x = week_end_date, y = y_hatcase_lm1), colour = "green")+
  geom_line(aes(x = week_end_date, y = y_hatcasegam), colour = "blue", size = 2)+
  geom_line(aes(x = week_end_date, y = case_sum))+
  scale_y_log10(labels = scales::label_log())+
  scale_x_date(labels = scales::label_date_short(),
               date_breaks = "2 week",
               name = "")+
  theme_classic()
```


## BORING LM revisit
Not the best fit but the closet to the shape of the data - few extra wiggles

```{r lm BORING reminder}
ggplot(data = National_weekly)+
  geom_line(aes(x = week_end_date, y = case_sum))+
  geom_line(aes(x = week_end_date, y = y_hatcase_lm), colour = "brown", size = 1.5)+
  scale_y_log10(labels = scales::label_log())+
  scale_x_date(labels = scales::label_date_short(),
               date_breaks = "2 week",
               name = "")+
  theme_classic()
```

### X weeks predicting the next week (or this week)

for predicting the fututrre, we are not gfoing to have the present day WW results, and the lag of the WW to cases is not enought o usefully model case loads.
An updated model might look like:  
`r case_sum ~ cases_last_week + copies_last_week + end_of_week_date`



```{r}
National_weekly$mod_test <- NA
train_length <- 5
for(i in 1:(length(National_weekly$x) - train_length)){
  
  # get data for training and testing
  data_train <- National_weekly %>%
    filter(x >= i, x < i + train_length)
  
  data_test <- National_weekly %>%
    filter(x == i+train_length)
  
  # train model for the following week
  mod_train <- lm(case_sum ~ cases_last_week + copies_last_week + x, data = data_train)
  
  # get prediction for follwoing week
  National_weekly$mod_test[i+train_length] <- predict(mod_train,
                                                      list(x = data_test$x,
                                                           copies_per_person_day = data_test$copies_per_person_day,
                                                           cases_last_week = data_test$cases_last_week,
                                                           copies_last_week = data_test$copies_last_week))
}

ggplot(data = National_weekly)+
  geom_line(aes(x = week_end_date, y = case_sum))+
  geom_line(aes(x = week_end_date, y = y_hatcase_lm), colour = "brown")+
  geom_line(aes(x = week_end_date, y = mod_test), colour = "orange", size = 1.5)+
  scale_y_log10(labels = scales::label_log())+
  scale_x_date(labels = scales::label_date_short(),
               date_breaks = "2 week",
               name = "")+
  theme_classic()

National_weekly$y_hatcase_lm - National_weekly$mod_test
National_weekly$case_sum - National_weekly$mod_test

mean((National_weekly$case_sum - National_weekly$mod_test)^2, na.rm = T)^0.5
```


